---
title: "Local FDR"
author: "Bradley Efron, Brit B. Turnbull and Balasubramanian Narasimhan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Local FDR}
  %\VignetteKeywords{FDR, local FDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette includes locfdr's complete help documentation, including
usage tips, which could not fit in the R help file.  It also
demonstrates usage of locfdr through an example using the simulated
data included in the package.

## Description and Usage

locfdr computes local false discovery rates, following the definitions
and description in the references listed below.

```
locfdr(zz, bre=120, df=7, pct=0, pct0=1/4, nulltype=1, type=0, plot=1,
        mult, mlests, main=" ", sw=0)
```

## Arguments

### zz

zz is a vector of summary statistics, one for each case under
consideration.  In a microarray experiment, there would be one element
of zz for each gene, perhaps a $t$-statistic comparing gene expression
levels under two different conditions.  The calculations assume a large
number of cases, say `length(zz)` exceeding 200.

Results may be improved by transforming zz so that its elements are
theoretically distributed as $N(0,1)$ under the null hypothesis.  For
example, when using $t$-statistics, transform them by `zz =
qnorm(pt(t,df))`.  Recentering and rescaling zz may be necessary if
its central histogram looks very far removed from mean 0 and variance 1.

When using a permutation null distribution with sample zperm, transform
the original statistics zorig by `zz = qnorm(ecdf(zperm)(zorig))`.
Such transformation is especially important when the theoretical null
option is invoked (see nulltype below).

### bre

bre is the number of breaks in the discretization of the $z$-score axis,
or a vector of breakpoints fully describing the discretization.  If
`length(zz)` is small, such as when the number of cases is less than
about 1000, set bre to a number less than the default of 120.

### df

df is the degrees of freedom for fitting the estimated density $f(z)$
(see type below).  Larger values of df may be required if $f(z)$ has
sharp bends or other irregularities.  A warning is issued if the fitted
curve does not adequately match the histogram counts.  It is a good idea
to use the plot option to view the histogram and fitted curve.

### pct

pct is the excluded tail proportions of zz's when fitting $f$.  The
default `pct=0` includes the full range of zz's.  pct can also be a
2-vector, describing the fitting range.

### pct0

pct0 is the proportion of the zz distribution used in fitting the null
density $f_0$ by central matching.  If it is a 2-vector, e.g.
`pct0=c(0.25,0.60)`, the range `[pct0[1], pct0[2]]` is used.  If a
scalar, `[pct0, 1-pct0]` is used.

### nulltype

nulltype is the type of null hypothesis assumed in estimating $f_0$, for
use in the fdr calculations.

- 0 is the theoretical null $N(0,1)$, which assumes that zz has been
  scaled to have a $N(0,1)$ distribution under the null hypothesis.
- 1 (the default) is the empirical null with parameters estimated by
  maximum likelihood.
- 2 is the empirical null with parameters estimated by central matching.
- 3 is a "split normal" version of 2, in which $f_0(z)$ is allowed to
  have different scales on the two sides of the maximum.

Unless sw is set to 2 or 3, the theoretical, maximum likelihood, and
central matching estimates all will be output in the matrix fp0, and
both the theoretical and the specified nulltype will be used in the
calculations output in mat, but only the specified nulltype is used in
the calculation of the output fdr (local fdr estimates for every case).

### type

type is the type of fitting used for $f$.

- 0 is a natural spline.
- 1 is a polynomial.

In either case, $f$ is fit with degrees of freedom df (so total degrees
of freedom including the intercept is $\text{df}+1$).

### plot

plot specifies the plots desired.

- 0 gives no plots.
- 1 (the default) gives a single plot showing the histogram of zz and
  fitted mixture density $f$ (green solid curve) and null subdensity
  $p_0 f_0$ (blue dashed curve).  Colored histogram bars indicate
  estimated non-null counts.  Yellow triangles on the zz-axis indicate
  threshold values for $\text{fdr}(z) \leq 0.2$, if such cases exist.
- 2 also gives plot of fdr, and the right and left tail area Fdr curves.
- 3 gives instead the $f_1$ cdf of the estimated fdr curve.
- 4 gives all three plots.

We recommend setting plot to 1 or greater, to check the fit of $p_0
f_0$ to the histogram.  (If the fit is poor, try a different nulltype or
a different value of the mlests argument.)

### mult

mult is an optional scalar multiple (or vector of multiples) of the
sample size for calculation of the corresponding hypothetical Efdr
value(s).

### mlests

mlests is an optional vector of initial values for $(\delta_0,
\sigma_0)$ in the maximum likelihood iteration.  In addition, these are
used to determine the interval over which the maximum likelihood
estimation is performed.  If, for example, zz was transformed
quantile-wise from F statistics, most of zz's elements corresponding to
interesting features will be positive.  To shift the interval away from
such elements, specify a negative initial value for $\delta_0$, the
first element of mlests.  If the default results in a poor fit of $p_0
f_0$ to the histogram in the first plot, try setting mlests to move the
estimates toward the values suggested by the histogram.

### main

main is the main heading for the histogram plot.

### sw

sw determines the type of output desired.

- 2 gives a list consisting of the last 5 values listed under Value
  below.
- 3 gives the square matrix of dimension bre-1 representing the
  influence function of $\log(\text{fdr})$.  The $(i,j)$ entry of the
  matrix is the derivative of $\log(\text{fdr})$ at the midpoint of bin
  $i$ with respect to the count value of bin $j$.
- Any other value of sw returns a list consisting of the first 7 (8 if
  mult is supplied) values listed below.

## Value

### fdr

fdr is the estimated local false discovery rate for each case, using the
selected type and nulltype.

### fp0

fp0 is a matrix containing the estimated parameters delta (mean of
$f_0$), sigma (standard deviation of $f_0$), and p0 (proportion of
tests that are null), along with their estimated standard errors.  If
`nulltype < 3`, fp0 is a $5 \times 3$ matrix, with columns representing
delta, sigma, and p0 and rows representing nulltypes and estimate
vs. standard error.  If `nulltype == 3`, the second column corresponds to
the estimate of sigma for the left side of $f_0$, and a fourth column
corresponds to the sigma estimate for the right.

### Efdr

Efdr is the expected local false discovery rate for the non-null cases, a
measure of the experiment's power.  Large values of Efdr, say `Efdr >
0.4`, indicate low power.  Overall Efdr and right and left values are
given, both for the specified nulltype and for nulltype 0.  (If `nulltype
== 0`, values are given for nulltypes 1 and 0.)

### cdf1

cdf1 is a $99 \times 2$ matrix giving the estimated cdf of fdr under the
non-null distribution $f_1$.  Large values of the cdf for small fdr
values indicate good power.  Set plot to 3 or 4 to see the plot of cdf1.

### mat

mat is a $(\text{bre}-1) \times 11$ matrix, convenient for comparisons
and plotting.  Each row corresponds to a bin of the zz histogram, and
the columns contain the following:

1. **x**: the midpoint of the bin.
2. **fdr**: the estimated local false discovery rate at $x$.
3. **Fdrleft**: the left tail false discovery rate at $x$.
4. **Fdrright**: the right tail false discovery rate at $x$.
5. **f**: the mixture density estimate at $x$.
6. **f0**: the null density estimate at $x$.
7. **f0theo**: the null density estimate at $x$, using the theoretical
   null $N(0,1)$.
8. **fdrtheo**: the local false discovery rate at $x$, using `nulltype=0`.
9. **counts**: the number of elements of zz in the bin.
10. **lfdrse**: the delta-method estimate of the standard error of the
    log of the local false discovery rate.
11. **p1f1**: the estimated subdensity of the non-null zz elements.

### z.2

z.2 is the interval along the zz-axis outside of which $\text{fdr}(z) <
0.2$, the locations of the yellow triangles in the histogram plot.  If
no elements of zz on the left or right satisfy the criterion, the
corresponding element of z.2 is NA, and the corresponding triangle does
not appear.

### call

call is the function call.

### mult

If the argument mult was supplied, the value mult is the vector of the
ratios of the hypothetical Efdr for the supplied multiples of the sample
size to the Efdr for the actual sample size.

### pds

pds is the vector of estimates of p0, delta, and sigma.

### x

x is the vector of bin midpoints.

### f

f is the vector of estimated values of $f(x)$ at the bin midpoints.

### pds.

pds. is the derivative of the estimates of p0, delta, and sigma with
respect to the bin counts.

### stdev

stdev is the vector of delta-method estimates of the standard deviations
of the p0, delta, and sigma estimates.

## Simulation Example

This simulation example involves 2000 "genes", each of which has
yielded a test statistic $z_i$, with $z_i \sim N(\mu_i, 1)$,
independently for $i = 1, 2, \ldots, 2000$.

Here $\mu_i$ is the "true score" of gene $i$, which we observe only
noisily.  1800 (90%) of the $\mu_i$ values are zero; the remaining 200
(10%) are from a $N(3,1)$ distribution.  The data are contained in the
dataset `lfdrsim`, where the $z_i$ are the column `zex`.

```{r preliminaries}
library(locfdr)
data(lfdrsim)
zex <- lfdrsim[, 2]
```

If we are confident that the null $z_i$'s are distributed as $N(0,1)$,
we run `locfdr` with `nulltype=0`.  Otherwise, we use the default
`nulltype=1`, which uses empirical estimates of the null density
parameters.

```{r compute-local-fdr, fig.width=6, fig.height=5}
w <- locfdr(zex)
```

In the figure, the green solid line is the spline-based estimate of the
mixture density $f$.  The blue dashed line is the null subdensity $p_0
f_0$, estimated by default by maximum likelihood (nulltype=1).
Whichever nulltype is specified, `locfdr` returns a matrix `fp0`
containing parameters of all three nulltypes and corresponding estimates
of the proportion $p_0$ of cases that are null, along with standard
errors.  In this example, the null distribution is $N(0,1)$, and both
the MLE and central matching estimates come close to this.

```{r show-fp0}
w$fp0
```

The output mat contains estimates of the local false discovery rates and
other functions for each bin midpoint $x$.

```{r show-mat}
w$mat[1:5, ]
```

The output fdr contains the local false discovery rate estimate for each
$z_i$.  One might use this vector to create a list of interesting cases.

```{r list-interesting-cases}
which(w$fdr < .2)
```

Here $0.2$ is a rule-of-thumb cut-off.  In the simulated data, the
first 200 cases have nonzero $\mu_i$.  So we can find the observed tail
false discovery proportion.

```{r tail-fdp}
sum(which(w$fdr < .2) > 200) / sum(w$fdr < .2)
```

The estimated tail FDR can be found from the `mat` output.

```{r tail-fdr}
w$mat[which(w$mat[, "fdr"] < .2)[1], "Fdrright"]
```

The tail FDR is the mean local fdr over the entire tail and is therefore
smaller than the local fdr cutoff.

## References

1. Efron, B. (2004) "Large-scale simultaneous hypothesis testing: the
   choice of a null hypothesis," *JASA*, **99**, pp. 96--104.

2. Efron, B. (2005) "Local False Discovery Rates."

3. Efron, B. (2006) "Size, Power, and False Discovery Rates."

4. Efron, B. (2006) "Correlation and Large-Scale Simultaneous
   Significance Testing," *JASA*, **102**, pp. 93--103.
